---
title: "Evidence Interpretation and Methodological Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evidence Interpretation and Methodological Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(SEMANTICA))
```

# Purpose

SEMANTICA's sample-free workflow analyzes the language of psychometric items.
It can screen semantic construct alignment, overlap, redundancy, required-facet
coverage, wording polarity, and compatibility with a proposed structural
blueprint. These outputs are **semantic-proxy diagnostics**. They are not a
replacement for participant-based psychometric evidence.

A defensible development program therefore separates four questions:

1. **Semantic signal:** Do item embeddings contain information about human item
   relations?
2. **Calibration:** Can that signal predict participant correlations across
   instruments, constructs, models, and languages?
3. **Selection:** Does SEMANTICA choose better item sets than simpler methods
   under the same computational budget?
4. **Confirmation:** Do SEMANTICA-assisted forms perform well in independent
   participant samples?

This vignette documents the APIs added to make those questions testable.

# Evidence status

```{r evidence-example}
mock <- structure(
  list(
    construct_coverage = NULL,
    participant_validation_performed = FALSE,
    optimization = list(response_validation = NULL)
  ),
  class = c("semantica_full_pipeline_result", "list")
)
semantica_evidence_status(mock)
```

Do not describe semantic PFA/ESEM/DFI as establishing reliability, construct
validity, invariance, DIF, or criterion validity. When response data are
supplied, report the response-data analysis separately.

# Selection-conditioned and local diagnostics

SEMANTICA 0.2.10 keeps semantic discrimination interpretable across the item-selection boundary. The threshold-free `A` statistic is a stochastic-superiority effect-size descriptor (Vargha & Delaney, 2000), but an optimizer deliberately searches many candidate subsets. General post-selection theory shows why statistics computed after data-driven selection should not inherit naive pre-selection inferential interpretations (Berk et al., 2013). SEMANTICA therefore treats selected `A` as **post-selection descriptive evidence**, not as selection-adjusted inference.

`semantica_selection_context()` reports the candidate-pool value, selected-set value, their descriptive change, and factor-specific diagnostics together. No universal post-selection cutoff is introduced:

```{r selection-context, eval=FALSE}
ctx <- semantica_selection_context(
  wrapped$cosine_sim_matrix,
  stats::setNames(wrapped$df$type, wrapped$df$item),
  result$best_items
)
ctx$pool$estimate
ctx$selected$estimate
ctx$stochastic_superiority_gain
ctx$selected_factor_diagnostics
```

`semantica_factor_semantic_diagnostics()` is deliberately local. A favorable aggregate can coexist with one weak factor, so each intended factor receives its own within-vs-between mean gap and stochastic-superiority `A`. This follows the broader SEM principle that global or aggregate behavior should not substitute for diagnosing localized misspecification (Saris, Satorra, & van der Veld, 2009). The function adds no new psychometric threshold.

PFA reporting now gives greater prominence to `partition_agreement_ari`. The adjusted Rand index explicitly corrects partition agreement for agreement expected by chance (Hubert & Arabie, 1985; Warrens & van der Hoef, 2022). `factor_presence_recovery` remains available because it answers a different question--whether intended factors are represented among dominant PFA dimensions--but it must not be interpreted as item-level partition accuracy.

ESEM technical admissibility remains separate from structural quality. ESEM was developed to permit cross-loadings rather than force an independent-clusters CFA structure (Asparouhov & Muthen, 2009), so convergence/admissibility alone is not substantive evidence that the intended factor partition is clean. SEMANTICA now makes `admissible_but_structurally_mixed` visually explicit and aggregates the already-computed item-level ESEM diagnostics by factor. HTMT-like overlap remains complementary evidence; discriminant-separation diagnostics can detect problems that simpler loading inspections may miss (Henseler, Ringle, & Sarstedt, 2015).

Finally, `best_objective` is now accompanied by `objective_context`. It is an **optimization utility**, not a universal scale-quality score. Objective magnitudes are only conditionally comparable when the objective definition, weights, candidate-pool construction, and evidence regime are held fixed. In particular, a semantic/PFA fallback objective should not be compared substantively with an ESEM-guided objective from another run.

References used for these interpretation rules:

- Asparouhov, T., & Muthen, B. (2009). Exploratory structural equation modeling. *Structural Equation Modeling, 16*(3), 397-438. https://doi.org/10.1080/10705510903008204
- Berk, R., Brown, L., Buja, A., Zhang, K., & Zhao, L. (2013). Valid post-selection inference. *The Annals of Statistics, 41*(2), 802-837. https://doi.org/10.1214/12-AOS1077
- Henseler, J., Ringle, C. M., & Sarstedt, M. (2015). A new criterion for assessing discriminant validity in variance-based structural equation modeling. *Journal of the Academy of Marketing Science, 43*, 115-135. https://doi.org/10.1007/s11747-014-0403-8
- Hubert, L., & Arabie, P. (1985). Comparing partitions. *Journal of Classification, 2*, 193-218.
- Saris, W. E., Satorra, A., & van der Veld, W. M. (2009). Testing structural equation models or detection of misspecifications? *Structural Equation Modeling, 16*(4), 561-582. https://doi.org/10.1080/10705510903203433
- Vargha, A., & Delaney, H. D. (2000). A critique and improvement of the CL common language effect size statistics of McGraw and Wong. *Journal of Educational and Behavioral Statistics, 25*(2), 101-132. https://doi.org/10.3102/10769986025002101
- Warrens, M. J., & van der Hoef, H. (2022). Understanding the adjusted Rand index and other partition comparison indices based on counting object pairs. *Journal of Classification, 39*, 487-509. https://doi.org/10.1007/s00357-022-09413-z

# Representation policy and content-alignment safeguards

SEMANTICA 0.2.7 separates the embedding model's task instructions from the
psychometric decision rules. `embedding_task = "auto"` applies a model-specific
instruction only for embedding families whose documented interface requires one;
unknown models are not modified. The resolved instruction policy is retained in the
result so two analyses that used different task prompts are not treated as equivalent.

Content alignment uses factor definitions as semantic prototypes. The assigned-vs-
competing factor margin is always retained as a diagnostic. In 0.2.7, automatic
exclusion is **not** the default: `content_alignment_mode = "diagnostic"` reports the
evidence without restricting candidates. If an analyst explicitly requests
`content_alignment_mode = "guard"`, a negative top-rank margin is called a clear
mismatch only when it exceeds a pool-relative reference derived from positively
separated items assigned to the same factor. If that reference cannot be estimated,
the case remains ambiguous rather than being automatically excluded. Explicit
forbidden/excluded concepts are also treated contrastively: an exclusion conflict is
flagged only when an excluded concept is closer than the assigned construct
definition. The guard remains feasibility-aware and never makes the requested
per-factor item count impossible. Facet ambiguity is diagnostic/coverage evidence,
not an automatic item-deletion rule.

Required-facet coverage now distinguishes metadata coverage from semantic-definition
coverage when facet descriptions and alignment diagnostics are available. A label
alone therefore no longer counts as semantic evidence that the wording represents a
facet.

Polarity remains a separate diagnostic by default. The built-in rule-based screen can
identify overt negation, but overt negation is not synonymous with reverse-keying.
Users whose item-writing protocol forbids such wording can opt into the feasibility-
aware polarity guard.

# Construct blueprints

```{r blueprint}
factors <- list(
  Clarity = list(
    description = "Clear and organized thinking.",
    facets = list(focus = list(), organization = list()),
    forbidden = c("memory capacity")
  ),
  Flexibility = list(
    description = "Adaptation of thinking when demands change.",
    facets = list(adaptation = list(), perspective_shift = list())
  )
)

expected <- matrix(
  c(1, .25,
    .25, 1),
  2, 2,
  dimnames = list(names(factors), names(factors))
)

blueprint <- semantica_construct_blueprint(
  factors,
  expected_relations = expected
)
```

`expected_relations` is an expected **semantic similarity pattern**. It must
not be copied from an empirical latent-factor correlation matrix and then
interpreted as though the two metrics were equivalent.

The optional ACO `nomological_weight` defaults to zero. A non-zero value should
be treated as a research hypothesis and externally calibrated.

# Pool-relative threshold calibration

Raw cosine values are not necessarily comparable across embedding models.
`semantica_calibrate_similarity_thresholds()` derives an experimental reference
from the current pool:

```{r threshold-example}
ids <- paste0("i", 1:6)
M <- matrix(c(
  1,.70,.62,.20,.15,.18,
  .70,1,.66,.18,.12,.20,
  .62,.66,1,.22,.16,.19,
  .20,.18,.22,1,.72,.68,
  .15,.12,.16,.72,1,.70,
  .18,.20,.19,.68,.70,1
), 6, 6, byrow = TRUE, dimnames = list(ids, ids))
fa <- setNames(c(rep("A", 3), rep("B", 3)), ids)
semantica_calibrate_similarity_thresholds(M, fa)
```

The returned object is deliberately labeled an experimental heuristic. Use it
for sensitivity analysis until it has been benchmarked against expert and
participant outcomes.

# Semantic-to-response calibration

Given an empirical item-correlation matrix from participant data:

```{r calibration-example}
R <- M
R[lower.tri(R)] <- .05 + .75 * M[lower.tri(M)]
R[upper.tri(R)] <- t(R)[upper.tri(R)]
diag(R) <- 1

cal <- semantica_fit_empirical_calibration(
  semantic_matrix = M,
  response_matrix = R,
  method = "fisher_linear"
)
cal$diagnostics
```

A fitted mapping is a **training result**, not external validation. Apply it to
new matrices only when the training and target settings are theoretically
comparable:

```{r apply-calibration}
calibrated <- semantica_apply_empirical_calibration(M, cal)
```

# Leave whole scales out

Randomly splitting item pairs leaks instrument-specific wording and construct
context between training and test sets. Prefer whole-instrument holdout when
possible:

```{r loso}
case1 <- list(semantic_matrix = M, response_matrix = R)
case2 <- list(semantic_matrix = M * .95 + diag(.05, 6), response_matrix = R)

cv <- semantica_leave_one_scale_out_calibration(
  list(scale_1 = case1, scale_2 = case2),
  method = "linear"
)
cv$per_scale
```

For publication-grade benchmarking, include many independent instruments and
report uncertainty across held-out scales rather than only pooled item-pair
metrics.

# Polarity and signed relations

Cosine similarity is primarily a measure of semantic relatedness and can remain
high when wording direction changes. SEMANTICA therefore keeps polarity as a
separate evidence dimension:

```{r polarity}
items <- data.frame(
  item_id = c("a", "b"),
  item_text = c(
    "I feel capable when challenges arise.",
    "I do not feel capable when challenges arise."
  )
)
semantica_polarity_diagnostics(items)
```

The built-in detector is deliberately conservative. A separately benchmarked
NLI/relation scorer can be supplied by advanced users. The experimental
`semantica_signed_semantic_matrix()` requires analyst-supplied polarity and
therefore does not claim to infer item direction automatically.

# Multi-model robustness

Conclusions should ideally survive plausible representation choices:

```{r robustness}
M2 <- M
M2[1, 2] <- M2[2, 1] <- .64
rob <- semantica_semantic_robustness(list(model_A = M, model_B = M2))
rob$matrix_rank_agreement

ensemble <- semantica_ensemble_similarity(
  list(model_A = M, model_B = M2),
  method = "rank_mean"
)
```

`rank_mean` deliberately emphasizes pair ordering rather than assuming that
raw cosine values from different models share a common metric scale.

# Matrix-repair reporting

Proxy ESEM needs a positive-definite correlation-like matrix. Any transformation
changes the original semantic geometry. Report that change:

```{r repair, eval=FALSE}
repair <- semantica_matrix_repair_diagnostics(
  cosine_matrix = M,
  factor_assignment = fa,
  factors = c("A", "B")
)
repair[c("min_eigen_before", "min_eigen_after", "frobenius_change", "max_abs_change")]
```

A heavily repaired matrix should be treated as a warning, not as a successfully
"fixed" source of stronger evidence.

# Equal-budget optimizer ablations

The complexity of ACO should be justified empirically. Use
`semantica_compare_selection_methods()` with random, greedy, semantic-only ACO,
and full SEMANTICA methods under a common budget. The evaluator should use an
independent criterion, ideally held-out participant data.

# Counterfactual stress tests

Controlled modifications are useful methodological unit tests. For example, create
paraphrases, negations, neighboring-construct substitutions, and unrelated
items, then verify that diagnostics move in theoretically predicted directions.
`semantica_counterfactual_stress_test()` provides a generic harness while
leaving the scoring function explicit.

# Pilot updating

`semantica_update_with_pilot()` can learn a semantic-to-response calibration
from pilot data. The pilot sample should not then be reused as the confirmatory
sample for final performance claims. A clean workflow is:

**semantic design -> pilot calibration/item revision -> frozen scale -> new
confirmatory participant sample**.

# Recommended benchmark reporting

At minimum report:

- exact embedding model/provider/revision when available;
- preprocessing and normalization;
- semantic matrix method;
- matrix-repair magnitude;
- calibration training instruments;
- whole-instrument holdout performance;
- embedding-model robustness;
- optimizer baseline comparisons under equal budgets;
- participant sample and empirical correlation estimator;
- all SEMANTICA and R/package versions;
- random seeds and computational budget.

Use `semantica_save_bundle()` to retain these artifacts in a sanitized,
versioned result bundle.

## Additional experimental research scaffolds

The following APIs implement hypotheses from the SEMANTICA research roadmap so
they can be benchmarked explicitly. They are not production validity rules.

### Contrastive construct prototypes

```{r, eval=FALSE}
alignment <- semantica_contrastive_construct_alignment(
  item_embeddings = item_embeddings,
  factor_assignment = factor_assignment,
  construct_embeddings = construct_prototypes
)
```

A positive contrastive margin says that the *semantic representation* of an
item is closer to its intended prototype than to the strongest supplied
competitor. It is not empirical discriminant validity.

### Semantic fingerprints and scale-version comparison

```{r, eval=FALSE}
fp <- semantica_scale_fingerprint(cosine_matrix, factor_assignment)
comparison <- semantica_compare_scale_versions(cosine_v1, cosine_v2)
```

These functions are descriptive tools for instrument geometry and revision
tracking. They do not replace longitudinal validation.

### Cross-language semantic screening

```{r, eval=FALSE}
translation_screen <- semantica_cross_language_equivalence(
  reference_embeddings,
  translated_embeddings
)
```

This requires a shared multilingual embedding space. High paired similarity or
neighborhood preservation may support translation review, but neither is
measurement invariance, DIF evidence, or proof of cross-cultural equivalence.

### Human adjudication queue

```{r, eval=FALSE}
queue <- semantica_expert_review_queue(
  matrices = list(model_a = cosine_a, model_b = cosine_b),
  polarity = polarity_diagnostics
)
```

Model disagreement is treated as a reason to ask an expert, not as an automatic
item-rejection rule.

### Coverage drafting and preregistration manifests

```{r, eval=FALSE}
prompts <- semantica_suggest_coverage_prompts(coverage, blueprint)
manifest <- semantica_preregistration_manifest(config, inputs = items)
```

Coverage prompts support item drafting only. The manifest freezes configuration
and input hashes for reproducibility; it does not convert a semantic screening
plan into participant-based validity evidence.
