---
title: "SEMANTICA Function Reference by Task"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMANTICA Function Reference by Task}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Main workflow

Start here:

- `semantica_run()` — scale-development workflow.
- `semantica_run_plan()` — preview counts, facets, workload, backend/model choices.
- `semantica_check_setup()` — pre-run backend/credential/model readiness.
- `semantica_list_backends()` — user-oriented backend registry.
- `semantica_run()` returns a six-part surface (`scale`, `items`, `diagnostics`, `plots`, `provenance`, `advanced`); `plots` retains the core report graphics and `advanced` retains the canonical result.
- `semantica_view()` — grouped navigation for the canonical result; compact or advanced views are available for either interface.
- `semantica_items()` — final selected scale.
- `semantica_overview()` — legacy plain-language overview helper.
- `summary()` — detailed diagnostic report.
- `plot()` — summary visualization; the regular result also retains the core report plots under `$plots`.
- `semantica_evidence_status()` — evidence-family status.
- `semantica_export()` — human-readable selected-scale export.
- `semantica_save_bundle()` / `semantica_load_bundle()` — analysis persistence.
- `semantica_cache_info()` / `semantica_clear_cache()` — embedding-cache operations.

# Extended configuration

Use these when you need additional control:

- `semantica_full_pipeline()`
- `semantica_llm_config()`
- `semantica_item_count_config()`
- `semantica_generation_config()`
- `semantica_quality_config()`
- `semantica_compute_config()`
- `semantica_resource_config()`
- `semantica_pfa_config()`
- `semantica_esem_config()`
- `semantica_fit_calibration_config()`
- `semantica_diagnostics_config()`
- `semantica_plot_config()`
- `semantica_construct_blueprint()`

An explicitly supplied top-level ESEM search option in
`semantica_full_pipeline()` takes precedence over the same convenience value
stored inside `semantica_esem_config()`.

# Existing item pools

- `semantica_connect(..., purpose = "embed")`
- `semantica_embed()`
- `semantica_import_embeddings()` — preferred boundary for external embeddings.
- `semantica_wrap()`
- `ACO_with_ESEM()` — component optimizer for prepared existing-item workflows.

Use the standard input columns `item_id`, `factor`, optional `facet`, and
`item_text`. See the user-workflows vignette for a complete recipe.

# Resources and recovery

- `semantica_capabilities()`
- `semantica_resource_plan()`
- `semantica_reset_resources()`
- `semantica_embedding_cache_config()`
- `semantica_setup_conda()`
- `semantica_activate_conda()`

For routine runs, `workers` is usually sufficient.

# Pipeline components

Use these to build or inspect individual stages:

- `semantica_connect()` / `semantica_backend_preflight()`
- `semantica_generate_items()` / `semantica_embed()` / `semantica_wrap()`
- `semantica_pipeline()`
- `ACO_with_ESEM()`
- `semantica_full_pipeline_custom()`
- `run_multi_seed_semantica()`
- archive/solution-history inspection functions

# Validation and research tools

Calibration, robustness, construct-graph, cross-language, counterfactual,
ensemble, empirical-calibration, and benchmark functions support validation and
methodological studies. See `vignette("evidence-interpretation", package =
"SEMANTICA")` for interpretation guidance.


# Result inspection and scripted workflows

High-level results support compact printing, detailed section-filtered summaries, and stable accessors:

```{r eval=FALSE}
result
summary(result, sections = c("aco", "structural"))
semantica_item_review(result)
semantica_factor_review(result)
semantica_diagnostics(result, "representation")
semantica_provenance(result)
semantica_config(result, changes_only = TRUE)
```

A run plan can be inspected and then executed with `semantica_execute()`. Configuration objects have a redacted print method so credential fields are not echoed to the console.

Participant responses collected after selection can be attached with `semantica_validate()`; this uses the stored selected scale and existing response-validation path rather than rerunning generation or ACO.
