---
title: "SEMANTICA Function Reference and Argument Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMANTICA Function Reference and Argument Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose and Scope

SEMANTICA supports early-stage scale construction by combining generated or
existing item text, semantic embeddings, ant colony optimization (ACO), and
sample-free ESEM/PFA diagnostics. Semantic and proxy-ESEM results assist
content screening; they are not substitutes for fitting measurement models to
item-response data in a suitable validation sample.

This vignette documents the supported public API. Dot-prefixed functions and
other unexported helpers are implementation details and should not be called
from analysis scripts.

## Offline Preparation Example

The preparation, export, and reload utilities can be learned without an API
key. This object is also used in the optimization examples below.

```{r offline-preparation}
library(SEMANTICA)

items_tbl <- data.frame(
  item_id = paste0("item_", 1:8),
  factor = rep(c("Clarity", "Flexibility"), each = 4),
  facet = rep(c("Expression", "Adaptation"), each = 4),
  item_text = c(
    "I organize my thoughts before I speak.",
    "I can express a complex idea clearly.",
    "I keep my explanations focused.",
    "I notice when my reasoning loses clarity.",
    "I adapt my thinking when evidence changes.",
    "I can switch strategies when needed.",
    "I consider a different approach when stuck.",
    "I adjust plans when conditions change."
  )
)

embeddings <- rbind(
  c(1.00, 0.10, 0.06), c(0.98, 0.13, 0.05),
  c(0.96, 0.17, 0.08), c(0.94, 0.19, 0.09),
  c(0.09, 1.00, 0.08), c(0.12, 0.98, 0.06),
  c(0.15, 0.96, 0.07), c(0.18, 0.94, 0.10)
)
rownames(embeddings) <- items_tbl$item_id

standard_items <- semantica_standardize_item_metadata(
  items_tbl,
  id_col = "item_id",
  dimension_col = "factor",
  facet_col = "facet",
  item_col = "item_text",
  strict = TRUE
)

embedded <- list(
  embeddings = embeddings,
  items_tbl = items_tbl,
  embed_model = "manual-example",
  embedding_diagnostics = list(normalized = FALSE)
)

wrapped <- semantica_wrap(
  embed_result = embedded,
  items_tbl = items_tbl,
  id_col = "item_id",
  factor_col = "factor",
  min_items_per_factor = 3L,
  cosine_adjustment = "none",
  semantic_calibration = NULL,
  compute_cosine_sensitivity = TRUE,
  verbose = FALSE
)
semantica_print_items(items_tbl, max_chars = 60L)

prefix <- file.path(tempdir(), "cognitive_agility")
semantica_export(wrapped, prefix = prefix)
reloaded <- semantica_reload(
  prefix = prefix,
  i.per.f = c(Clarity = 3L, Flexibility = 3L),
  default_i_per_f = 3L
)
```

## Backend and Session Functions

`SEMANTICA_BACKENDS` is a named registry rather than a function. It describes
the supported generation and embedding protocols.

| Function or object | Arguments | What it does |
|---|---|---|
| `SEMANTICA_BACKENDS` | None | Stores backend labels, endpoints, models, embedding support, and authentication settings. |
| `semantica_list_backends()` | None | Prints and invisibly returns `SEMANTICA_BACKENDS`. |
| `semantica_setup_conda()` | `env_name`, `conda`, `python_ver`, `packages`, `force`, `verbose` | Installs a Conda environment and optional extra Python packages for local Python backends. |
| `semantica_activate_conda()` | `env_name`, `conda`, `verbose` | Selects an existing Conda environment for the R session. |
| `semantica_connect()` | `backend`, `api_key`, `chat_model`, `embed_model`, `base_url`, `gguf_path`, `hf_token`, `timeout_s`, `verbose` | Creates a `semantica_session` for a cloud API, local server, or Python model. |
| `print.semantica_session()` | `x`, `...` | Prints the backend and chat model for a session; `...` is accepted for S3 compatibility. |

```{r sessions}
names(SEMANTICA_BACKENDS)
registry <- semantica_list_backends()

# Network/local environment actions are intentional user actions:
semantica_setup_conda(
  env_name = "semantica",
  conda = NULL,
  python_ver = "3.11",
  packages = character(0L),
  force = FALSE,
  verbose = TRUE
)
semantica_activate_conda(env_name = "semantica", conda = NULL, verbose = TRUE)

session <- semantica_connect(
  backend = "openai",
  api_key = Sys.getenv("OPENAI_API_KEY"),
  chat_model = NULL,
  embed_model = "text-embedding-3-small",
  base_url = NULL,
  gguf_path = NULL,
  hf_token = NULL,
  timeout_s = 120L,
  verbose = TRUE
)
print(session)
```

Use `base_url` with an OpenAI-compatible local server; use `gguf_path` for the
`python_llamacpp` backend; use `hf_token` only for gated Hugging Face models.

## Generation, Embedding, and Input Preparation

| Function | Arguments | Result |
|---|---|---|
| `semantica_standardize_item_metadata()` | `x`, `id_col`, `dimension_col`, `facet_col`, `item_col`, `strict` | A canonical table with `ID`, `Dimension`, `Facet`, and `item`. |
| `semantica_generate_items()` | `session`, `scale_name`, `scale_description`, `factors`, `response_format`, `item_style`, `language`, `n_per_factor`, `n_per_factor_override`, `overgenerate`, `max_retries`, `global_forbidden_max`, `temperature`, `verbose` | Candidate item text and item metadata. |
| `semantica_embed()` | `items_tbl`, `session`, `embed_session`, `text_col`, `id_col`, `batch_size`, `normalize`, `verbose` | Dense embeddings plus embedding diagnostics. |
| `semantica_wrap()` | `embed_result`, `items_tbl`, `id_col`, `factor_col`, `min_items_per_factor`, `cosine_adjustment`, `semantic_calibration`, `compute_cosine_sensitivity`, `verbose` | A cosine matrix, optimizer metadata, factor targets, and diagnostics. |
| `semantica_print_items()` | `items_tbl`, `max_chars` | Console display of generated items grouped by factor. |
| `semantica_export()` | `pipeline_result`, `prefix` | Three CSV files for items, optimizer metadata, and cosine similarities. |
| `semantica_reload()` | `prefix`, `i.per.f`, `default_i_per_f` | Reloaded optimizer input list from those CSV files. |

The `factors` argument can specify a simple dimension or facets within each
dimension. Setting `n_per_factor_override = TRUE` allocates
`n_per_factor` items across the declared facets rather than respecting nested
item counts.

```{r generation-and-embedding}
factors <- list(
  Clarity = list(
    description = "Clear and coherent expression of thought.",
    facets = list(
      Organization = list(description = "Structuring ideas logically."),
      Precision = list(description = "Expressing intended meaning precisely.")
    ),
    n_items = 12L
  ),
  Flexibility = list(
    description = "Adapting thought to changing evidence.",
    n_items = 12L
  )
)

items <- semantica_generate_items(
  session = session,
  scale_name = "Cognitive Agility",
  scale_description = "Clear and adaptive cognitive self-regulation.",
  factors = factors,
  response_format = "5-point Likert",
  item_style = "first-person declarative sentence",
  language = "English",
  n_per_factor = 12L,
  n_per_factor_override = TRUE,
  overgenerate = 2,
  max_retries = 3L,
  global_forbidden_max = 40L,
  temperature = 0.8,
  verbose = TRUE
)

embedded <- semantica_embed(
  items_tbl = items,
  session = session,
  embed_session = NULL,
  text_col = "item_text",
  id_col = "item_id",
  batch_size = 64L,
  normalize = TRUE,
  verbose = TRUE
)
```

## Preparation Pipeline

`semantica_pipeline()` runs connection, generation, embedding, and wrapping.
Arguments passed through `...` are generation controls such as
`response_format`, `item_style`, `language`, `overgenerate`, `max_retries`,
`global_forbidden_max`, and `temperature`.

| Argument | Meaning |
|---|---|
| `backend`, `embed_backend` | Generation and optional separate embedding backend. |
| `base_url`, `embed_base_url`, `api_key`, `embed_api_key`, `chat_model`, `embed_model`, `gguf_path` | Connection and model controls. |
| `embed_batch_size` | Number of texts included in each embedding request. |
| `scale_name`, `scale_description`, `factors`, `n_per_factor`, `n_per_factor_override` | Item-pool specification. |
| `cosine_adjustment`, `semantic_calibration`, `compute_cosine_sensitivity` | Semantic-proxy transformation and diagnostics. |
| `release_local_models`, `retain_embeddings` | Local-memory and retained-output controls. |
| `verbose`, `...` | Output and additional item-generation controls. |

```{r preparation-pipeline}
prepared <- semantica_pipeline(
  backend = "openai",
  embed_backend = NULL,
  base_url = NULL,
  embed_base_url = NULL,
  api_key = Sys.getenv("OPENAI_API_KEY"),
  embed_api_key = NULL,
  chat_model = NULL,
  embed_model = "text-embedding-3-small",
  embed_batch_size = 64L,
  gguf_path = NULL,
  scale_name = "Cognitive Agility",
  scale_description = "Clear and adaptive cognitive self-regulation.",
  factors = factors,
  n_per_factor = 12L,
  n_per_factor_override = TRUE,
  cosine_adjustment = "none",
  semantic_calibration = NULL,
  compute_cosine_sensitivity = TRUE,
  release_local_models = FALSE,
  retain_embeddings = TRUE,
  verbose = TRUE,
  temperature = 0.8
)
```

## ACO-ESEM Optimization

`ACO_with_ESEM()` selects an item subset from a prepared cosine matrix.
When `run_esem_during_search = TRUE`, solutions admitted to the ESEM-guided
archive must be scored through the requested ESEM pathway unless the user
explicitly chooses `esem_failure_policy = "semantic_fallback"`.

```{r optimization}
result <- ACO_with_ESEM(
  cosine_sim_matrix = wrapped$cosine_sim_matrix,
  df = wrapped$df,
  model_type = "correlated",
  i.per.f = c(Clarity = 3L, Flexibility = 3L),
  ants = 20L,
  max.iter = 10L,
  max_total_iter = NULL,
  max_esem_fits = NULL,
  esem_every = 5L,
  run_esem_during_search = FALSE,
  esem_weight = 0.50,
  esem_failure_policy = "stop",
  esem_sample_size = "auto",
  elite_k = 10L,
  dfi_mode = "heuristic_semantic",
  pfa_mode = "diagnostic",
  semantic_n_sensitivity = FALSE,
  final_dddfi = FALSE,
  final_equivtest = FALSE,
  history_mode = "summary",
  use_parallel = FALSE,
  verbose = FALSE
)
```

### Optimization Argument Catalogue

| Argument group | Arguments | Use |
|---|---|---|
| Inputs | `cosine_sim_matrix`, `df`, `model_type`, `i.per.f` | Semantic input matrix, item metadata, backward-compatible model label, and number of retained items per factor. |
| Search budget | `ants`, `max.iter`, `max_total_iter`, `archive_stable_window`, `elite_k` | Colony size, patience/iteration bounds, early stopping, and archive size. |
| Search-time ESEM | `esem_every`, `run_esem_during_search`, `max_esem_fits`, `esem_weight`, `esem_failure_policy`, `esem_eval_top_k`, `fast_esem`, `fast_esem_iter_max`, `full_esem_iter_max` | Frequency, computational budget, scoring weight, fallback behavior, and fit precision. |
| ESEM specification | `esem_sample_size`, `rotation`, `rotation_args`, `data_type`, `original_data`, `target_loadings`, `target_factor_cors` | Proxy sample size, rotation, observed-data type, and DFI target population settings. |
| Dynamic-fit calibration | `dfi_reps`, `dfi_level`, `dfi_criterion`, `dfi_mode`, `dfi_esem_reps`, `dfi_search_reps`, `dfi_roc_misspec_strength`, `dfi_esem_strategy`, `dfi_adaptive_min_reps`, `dfi_adaptive_batch_reps`, `dfi_adaptive_tol`, `dfi_adaptive_stable_batches`, `dfi_fallback_policy` | Simulation mode, replication budgets, adaptive stopping, and fallback order. |
| Final DFI diagnostics | `final_dfi_recalibrate`, `final_dfi_reps`, `final_dddfi`, `final_dddfi_reps`, `final_dddfi_mad_target`, `final_equivtest`, `loading_pattern`, `embed_reliability`, `residual_inflation`, `dfi_warmup_iters` | Optional final recalibration and companion approximate-fit diagnostics. |
| Semantic screening | `redundancy_threshold`, `dup_threshold`, `htmt_threshold`, `cohesion_quantile`, `cohesion_retention`, `within_similarity_target`, `within_similarity_band`, `facet_coverage_weight` | Redundancy control, content-pool retention, target similarity, and facet coverage. |
| Structural guard | `psychometric_guard_weight`, `psychometric_guard_min_ave`, `psychometric_guard_min_loading`, `psychometric_guard_min_primary_ge_50` | Soft penalties for weak convergent structure or improper solutions. |
| PFA diagnostics | `pfa_mode`, `pfa_weight`, `pfa_extraction`, `pfa_final_extraction`, `pfa_rotation`, `pfa_min_loading`, `pfa_min_margin` | Sample-free factor-structure diagnostic settings. |
| Reference-N checks | `reference_rmsea_close`, `reference_rmsea_poor`, `reference_power`, `reference_alpha`, `reference_max_n`, `semantic_n_sensitivity`, `semantic_n_grid`, `semantic_n_multipliers`, `semantic_n_iter_max` | Proxy reference sample-size calculation and sensitivity refits. |
| Scoring mode | `semantic_esem_score_mode`, `sigmoid_center`, `sigmoid_steepness`, `heuristic_beta`, `pheromone_update`, `fixed_evaporation`, `elite_pareto_rerank` | Objective weighting and ACO update behavior. |
| Validation planning | `validation_n_diagnostic`, `validation_n_reps`, `validation_n_grid`, `validation_n_max`, `validation_n_convergence`, `validation_n_max_heywood`, `validation_n_min_recovery`, `validation_n_max_loading_error`, `validation_n_min_dominance`, `validation_n_max_cross_error`, `validation_n_max_factor_cor_error`, `validation_data`, `validation_ordered` | Monte Carlo planning and optional final response-data validation fit. |
| Storage and computation | `debug_mode`, `keep_solution_history`, `history_mode`, `use_parallel`, `n.cores`, `verbose`, `...` | Runtime, retained history, and output. `n.cores` is capped at two parallel workers for resource-conscious execution. |

Use the semantic-only search in a quick dry run to examine item-pool behavior.
For substantive use, justify the ESEM/DFI settings, inspect convergence and
guard diagnostics, and validate the chosen indicators on response data.

## Multiple Seeds and Reports

| Function | Arguments | Use |
|---|---|---|
| `run_multi_seed_semantica()` | `seeds`, `cosine_sim_matrix`, `df`, `i.per.f`, `verbose_seeds`, `...` | Repeats `ACO_with_ESEM()` and reports selection stability; `...` supplies optimizer controls. |
| `report_semantica_v2()` | `result`, `digits` | Prints a compact model and optimization report. |
| `inspect_elite_archive()` | `result`, `top_n` | Prints top unique archived item selections. |
| `inspect_solution_history()` | `result`, `top_n`, `sort_by` | Prints retained search-history summaries ranked by an objective column. |

```{r reporting}
multi <- run_multi_seed_semantica(
  seeds = 1:3,
  cosine_sim_matrix = wrapped$cosine_sim_matrix,
  df = wrapped$df,
  i.per.f = c(Clarity = 3L, Flexibility = 3L),
  verbose_seeds = TRUE,
  ants = 20L,
  max.iter = 10L,
  run_esem_during_search = FALSE,
  dfi_mode = "heuristic_semantic",
  use_parallel = FALSE,
  verbose = FALSE
)

report_semantica_v2(result = result, digits = 3L)
inspect_elite_archive(result = result, top_n = 5L)
inspect_solution_history(result = result, top_n = 10L, sort_by = "total")
```

## Visualization API

All visualization functions consume an `ACO_with_ESEM()` result. Plots that
display item relationships also take the original cosine similarity matrix.

| Function | Arguments | Display |
|---|---|---|
| `plot_pheromone_heatmap()` | `result` | Pheromone trails and selected-item intensity. |
| `plot_fitness_evolution()` | `result` | Objective progression across iterations. |
| `plot_semantic_networks()` | `result`, `cosine_sim_matrix`, `edge_threshold_before`, `edge_threshold_after`, `df`, `max_before_items` | Before/after semantic networks. |
| `plot_esem_loadings()` | `result` | Rotated ESEM loading heatmap. |
| `plot_loading_profiles()` | `result` | Loading profiles with interval displays. |
| `plot_dfi_gauges()` | `result` | Fit measures compared with DFI cutoffs. |
| `plot_score_radar()` | `result` | Score-component radar chart. |
| `plot_semantic_discrimination()` | `result`, `cosine_sim_matrix` | Within- versus between-factor similarity. |
| `plot_interactive_semantic_space()` | `result`, `cosine_sim_matrix`, `label_all`, `mode`, `max_pool_items` | Interactive 2D or 3D MDS item map. |
| `plot_esem_path_diagrams()` | `result`, `cosine_sim_matrix`, `min_loading_show`, `show_crossloadings`, `df`, `min_factor_cor_show`, `max_factor_cor_edges`, `before_model`, `before_refit_max_items`, `before_proxy_max_items` | Before/after path diagrams using a proxy or an explicitly requested refit. |
| `plot_item_selection_frequency()` | `result`, `multi_result` | Item frequency in archive or repeated runs; supply at least one input. |
| `plot_crossloading_specificity()` | `result`, `ratio_cap`, `danger_ratio` | Cross-loading specificity diagnostic. |
| `plot_pfa_diagnostics()` | `result` | Sample-free PFA loading diagnostic. |
| `plot_semantic_n_sensitivity()` | `result` | Proxy reference-N sensitivity; may return `NULL` if not calculated. |
| `plot_summary_of_results()` | `result`, `cosine_sim_matrix` | Summary dashboard; cosine matrix is optional and enriches semantic panels. |
| `the_coolest_plot_ever()` | `result`, `cosine_sim_matrix`, `max_background_items`, `edge_quantile` | Presentation-oriented semantic distillation map. |

```{r individual-plots}
plot_pheromone_heatmap(result)
plot_fitness_evolution(result)
plot_semantic_networks(
  result, wrapped$cosine_sim_matrix,
  edge_threshold_before = 0.45,
  edge_threshold_after = 0.30,
  df = wrapped$df,
  max_before_items = 200L
)
plot_esem_loadings(result)
plot_loading_profiles(result)
plot_dfi_gauges(result)
plot_score_radar(result)
plot_semantic_discrimination(result, wrapped$cosine_sim_matrix)
plot_interactive_semantic_space(
  result, wrapped$cosine_sim_matrix,
  label_all = FALSE, mode = "2d", max_pool_items = 250L
)
paths <- plot_esem_path_diagrams(
  result, wrapped$cosine_sim_matrix,
  min_loading_show = 0.10,
  show_crossloadings = TRUE,
  df = wrapped$df,
  min_factor_cor_show = 0.05,
  max_factor_cor_edges = 15L,
  before_model = "proxy",
  before_refit_max_items = 60L,
  before_proxy_max_items = 150L
)
plot_item_selection_frequency(result = result, multi_result = multi)
plot_crossloading_specificity(result, ratio_cap = 10, danger_ratio = 1.5)
plot_pfa_diagnostics(result)
plot_semantic_n_sensitivity(result)
plot_summary_of_results(result, cosine_sim_matrix = wrapped$cosine_sim_matrix)
the_coolest_plot_ever(
  result, cosine_sim_matrix = wrapped$cosine_sim_matrix,
  max_background_items = 260L, edge_quantile = 0.65
)
```

`semantica_plot_all()` orchestrates the regular diagnostic set. It does not
call `the_coolest_plot_ever()` because that plot is intended as an optional
presentation aid.

| Argument | Meaning |
|---|---|
| `result`, `cosine_sim_matrix`, `df`, `multi_result` | Optimizer and optional repeated-run inputs. |
| `interactive_mode`, `include_interactive` | Type and inclusion of interactive MDS output. |
| `before_path_model`, `before_path_refit_max_items`, `network_max_items`, `mds_max_items`, `path_proxy_max_items` | Performance and path-display bounds. |
| `save`, `out_dir`, `device`, `width`, `height`, `dpi` | Plot file output options. |
| `progress` | Per-plot progress messages. |

```{r plot-wrapper}
plots <- semantica_plot_all(
  result = result,
  cosine_sim_matrix = wrapped$cosine_sim_matrix,
  df = wrapped$df,
  multi_result = multi,
  interactive_mode = "2d",
  save = FALSE,
  out_dir = "semantica_plots",
  device = "png",
  width = 12,
  height = 8,
  dpi = 180,
  before_path_model = "proxy",
  before_path_refit_max_items = 60L,
  network_max_items = 200L,
  mds_max_items = 250L,
  path_proxy_max_items = 150L,
  include_interactive = TRUE,
  progress = TRUE
)
```

## One-Call Workflow

`semantica_full_pipeline()` combines generation, embedding, wrapping,
optimization, reporting outputs, and optional visualization. It accepts:

| Argument group | Arguments |
|---|---|
| Scale specification | `scale_name`, `scale_description`, `factors`, `backend`, `candidate_items_per_factor`, `items_per_factor` |
| Generation and embedding | `generation_options = list(embed_backend, api_key, embed_api_key, chat_model, embed_model, embed_batch_size, base_url, gguf_path, cosine_adjustment, semantic_calibration, compute_cosine_sensitivity, release_local_models, retain_embeddings, ...)` |
| ACO-ESEM search | `optimization_options = list(ants, max.iter, esem_every, run_esem_during_search, esem_weight, rotation, elite_k, history_mode, use_parallel, n.cores, ...)` |
| DFI and semantic proxy calibration | `dfi_options = list(dfi_mode, dfi_reps, dfi_esem_reps, final_dfi_recalibrate, final_dddfi, reference_rmsea_close, semantic_n_sensitivity, ...)` |
| PFA diagnostics | `pfa_options = list(pfa_mode, pfa_weight, pfa_extraction, pfa_rotation, pfa_unit_diagnostics, ...)` |
| Validation planning | `validation_options = list(validation_n_diagnostic, validation_n_reps, validation_data, validation_ordered, ...)` |
| Plot generation | `plot_options = list(generate, interactive_mode, save, out_dir, device, width, height, dpi, include_interactive, ...)` |
| Runtime output | `verbose` |

For backward compatibility, legacy long-form names such as `n_per_factor`,
`i.per.f`, `dfi_mode`, and `generate_plots` can still be supplied directly by
name through `...`.

```{r full-workflow}
full <- semantica_full_pipeline(
  scale_name = "Cognitive Agility",
  scale_description = "Clear and adaptive cognitive self-regulation.",
  factors = factors,
  backend = "openai",
  candidate_items_per_factor = 12L,
  items_per_factor = c(Clarity = 4L, Flexibility = 4L),
  generation_options = list(
    api_key = Sys.getenv("OPENAI_API_KEY"),
    embed_model = "text-embedding-3-small",
    embed_batch_size = 64L,
    cosine_adjustment = "none",
    retain_embeddings = TRUE
  ),
  optimization_options = list(
    ants = 60L,
    max.iter = 30L,
    run_esem_during_search = TRUE,
    history_mode = "summary"
  ),
  dfi_options = list(dfi_mode = "auto"),
  pfa_options = list(pfa_unit_diagnostics = TRUE),
  plot_options = list(generate = TRUE, include_interactive = TRUE, save = FALSE),
  verbose = TRUE
)
```

## Interpretation and Reporting Boundaries

The semantic similarity matrix encodes wording/content proximity produced by
an embedding model. The ESEM and PFA diagnostics fitted to that proxy can help
find redundant, poorly differentiated, or structurally weak candidate item
sets. They do not establish factor structure in respondents. A final scale
requires response-data analyses, including appropriate ESEM/CFA estimation,
reliability and validity evidence, measurement invariance where relevant, and
replication.
