---
title: "SEMANTICA Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMANTICA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

SEMANTICA is designed for early-stage psychometric scale construction. It helps generate candidate items, represent item wording as embeddings, reduce semantic redundancy, and select item sets that are then evaluated through full exploratory structural equation modeling (ESEM).

## Define a Construct

```{r}
library(SEMANTICA)

factors <- list(
  Clarity = list(
    description = "Clear, organized, and coherent thinking.",
    n_items = 12
  ),
  Flexibility = list(
    description = "Ability to change cognitive strategy when demands change.",
    n_items = 12
  )
)
```

## Generate and Embed Items

These steps require API credentials or a local model backend.

```{r}
session <- semantica_connect("openai")

items <- semantica_generate_items(
  session = session,
  scale_name = "Cognitive Agility",
  scale_description = "Adaptive and clear cognitive self-regulation.",
  factors = factors,
  response_format = "5-point Likert",
  item_style = "first-person declarative sentence",
  language = "English"
)

embedded <- semantica_embed(items, session)
wrapped <- semantica_wrap(embedded)
```

## Optimize an Item Set

```{r}
result <- ACO_with_ESEM(
  cosine_sim_matrix = wrapped$cosine_sim_matrix,
  df = wrapped$df,
  i.per.f = c(Clarity = 4, Flexibility = 4),
  ants = 60,
  max.iter = 30,
  run_esem_during_search = TRUE,
  esem_sample_size = "auto",
  dfi_mode = "auto"
)

report_semantica_v2(result)
inspect_elite_archive(result, top_n = 5)
inspect_solution_history(result, top_n = 10)
```

## Visualize Diagnostics

```{r}
plots <- semantica_plot_all(
  result = result,
  cosine_sim_matrix = wrapped$cosine_sim_matrix,
  df = wrapped$df,
  save = FALSE
)

plots$p04_loadings_matrix
plots$p08_discrimination
plots$plot_summary_of_results
the_coolest_plot_ever(result, wrapped$cosine_sim_matrix)
```

Plot 10 shows BEFORE/AFTER ESEM path diagrams. When the full BEFORE-pool ESEM
cannot be refit, SEMANTICA falls back to a clearly labeled sample-free PFA or
semantic-proxy path view instead of returning a blank placeholder. Plot 10 also
draws factor-correlation rails when correlations are available. Plot 13
sign-anchors PFA dimensions so primary loadings are oriented positively for
the intended factor; negative cross-loadings can still appear and are
psychometrically interpretable. `the_coolest_plot_ever()` is a manual-only
semantic distillation map, so it is available for presentations without being
added to the default plot bundle.

## Export and Reload

```{r}
semantica_export(wrapped, prefix = "cognitive_agility")
reloaded <- semantica_reload("cognitive_agility")
```

The exported CSV files allow item text, item metadata, and cosine similarities to be reviewed or shared without rerunning generation and embedding.
