---
title: "SEMANTICA User Workflows: Existing Items, Embeddings, and Pilot Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMANTICA User Workflows: Existing Items, Embeddings, and Pilot Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Choose the workflow from your starting material

| You currently have | Start with |
|---|---|
| construct/factor definitions only | `semantica_run()` |
| item text but no embeddings | `semantica_embed()` -> `semantica_wrap()` |
| item text plus external embeddings | `semantica_import_embeddings()` -> `semantica_wrap()` |
| participant responses after a pilot | `semantica_full_pipeline(..., validation_data = ...)` |

# Item-table schema

For existing item pools, use:

```{r}
items <- data.frame(
  item_id = c("a1", "a2", "a3", "b1", "b2", "b3"),
  factor = rep(c("A", "B"), each = 3),
  facet = NA_character_,
  item_text = paste("Item", 1:6),
  stringsAsFactors = FALSE
)
```

`item_id` must be unique and stable. `factor` is the intended factor.
`facet` is optional. `item_text` is the wording to embed.

SEMANTICA accepts some historical aliases internally, but new user code should
prefer this schema.

# Existing text, no embeddings

Create only an embedding-purpose session:

```{r eval=FALSE}
embed_session <- semantica_connect(
  backend = "ollama",
  embed_model = "nomic-embed-text",
  purpose = "embed"
)

embedded <- semantica_embed(
  items_tbl = items,
  session = embed_session,
  verbose = FALSE
)

prepared <- semantica_wrap(
  embedded,
  verbose = FALSE
)
```

Then set the final number of items per factor and run the optimizer:

```{r eval=FALSE}
i_per_factor <- c(A = 3L, B = 3L)

selected <- ACO_with_ESEM(
  cosine_sim_matrix = prepared$cosine_sim_matrix,
  df = prepared$df,
  i.per.f = i_per_factor,
  pfa_mode = "diagnostic",
  dfi_mode = "heuristic_semantic",
  use_parallel = TRUE,
  seed = 20260828L
)
```

This workflow starts at the embedding stage because the item text already
exists.

# Existing external embeddings

Use `semantica_import_embeddings()` rather than manually fabricating an
embedding-result list:

```{r eval=FALSE}
imported <- semantica_import_embeddings(
  embeddings = embedding_matrix,
  items_tbl = items,
  embedding_ids = rownames(embedding_matrix),
  provider = "external",
  model = "my-embedding-model",
  normalize = FALSE
)

prepared <- semantica_wrap(imported, verbose = FALSE)
```

Embedding IDs and item IDs must match exactly; positional row alignment is not
assumed.

# Participant responses after a pilot

Participant data must be a data frame whose item columns use the same selected
item IDs. Extra non-item variables should be removed or handled outside the
validation input so item identity is unambiguous.

For ordinal items, supply the ordered item-column names through
`validation_ordered`.

```{r eval=FALSE}
result <- semantica_full_pipeline(
  scale_name = "Example Scale",
  scale_description = "...",
  factors = factors,
  llm = llm_cfg,
  chat_model = "...",
  embed_model = "...",
  validation_data = pilot_responses,
  validation_ordered = names(pilot_responses),
  seed = 20260828L
)
```

Then inspect:

```{r eval=FALSE}
semantica_evidence_status(result)
summary(result)
```

The participant-response fit is a separate evidence family. It does not turn
search-time embedding proxies into participant evidence.

If the scale was selected earlier and the result was preserved, participant data can be attached later without repeating generation or selection:

```{r eval=FALSE}
validated <- semantica_validate(
  result,
  responses = pilot_responses,
  ordered = names(pilot_responses)
)
summary(validated, sections = "participant")
```

The adapter requires the stored selected-item structure and fit-cutoff configuration from the original run.

# Saving an existing-item workflow

Prepared objects can be exported to CSV. For a complete high-level result,
`semantica_save_bundle()` preserves the full analysis provenance.
