---
title: "SEMANTICA: Recommended User Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMANTICA: Recommended User Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Start with `semantica_run()`

Start a new scale-development workflow with `semantica_run()`. When you need
settings beyond its arguments, use `semantica_full_pipeline()`. Both interfaces
use the same analysis pipeline.

A typical workflow is:

1. define the construct/factors;
2. check the generation/embedding setup;
3. preview item counts and expected workload;
4. run `semantica_run()`;
5. inspect selected items and evidence status;
6. plot and save;
7. collect participant responses and add participant-response validation when ready.

# 1. Define factors

Simple descriptions are sufficient:

```{r}
factors <- list(
  Goal_Direction = "Maintaining academic behavior around personally endorsed goals.",
  Monitoring = "Evaluating whether current study strategies are working."
)
```

Rich specifications can include `description`, `facets`, `forbidden`,
`examples`, and `extra_instructions`.

# 2. Check the backend

```{r eval=FALSE}
semantica_list_backends()
semantica_check_setup(llm = "ollama", probe = TRUE)
```

For cloud providers, credentials should normally be stored in `.Renviron`.
Anthropic and Groq are generation-only in the built-in registry, so pair them
with an embedding backend through `semantica_llm_config()`.

# 3. Preview the run

```{r eval=FALSE}
plan <- semantica_run_plan(
  scale_name = "Academic Self-Regulation Scale",
  scale_description = "Goal direction and monitoring in academic work.",
  factors = factors,
  pool_items = 12,
  selected_items = 4,
  overgenerate = 2,
  llm = "ollama"
)
plan
```

`pool_items = 12` means 12 retained candidates **per factor**. If a factor has
facets, those 12 are allocated across its facets. The plan also shows the
approximate raw generation target implied by `overgenerate`.

# 4. Run

```{r eval=FALSE}
result <- semantica_run(
  scale_name = "Academic Self-Regulation Scale",
  scale_description = "Goal direction and monitoring in academic work.",
  factors = factors,
  pool_items = 12,
  selected_items = 4,
  llm = list(
    backend = "ollama",
    chat_model = "llama3.2",
    embed_model = "nomic-embed-text"
  ),
  language = "English",
  response_format = "5-point Likert",
  item_style = "brief first-person declarative sentence",
  aco = "standard",
  seed = 20260828L
)
```

The ACO presets control search effort and diagnostic cadence:

- `fast`: lower effort for iteration/first-pass work;
- `standard`: routine scale-development runs;
- `full`: more intensive structural checking/calibration.


By default, console output is concise. Use `progress = "detailed"` when you need
component telemetry and `progress = "quiet"` in scripts.

# 5. Inspect the scale and evidence

```{r eval=FALSE}
result
result$scale
result$items
result$diagnostics
result$plots
result$provenance
result$advanced
summary(result)
summary(result, sections = c("semantic", "structural"))
semantica_evidence_status(result, labels = "both")
semantica_models(result)
semantica_config(result, "generation")
semantica_config(result, changes_only = TRUE)
semantica_provenance(result)
semantica_item_review(result)
semantica_factor_review(result)
```

`semantica_run()` returns a six-part object surface: `scale`, `items`, `diagnostics`, `plots`, `provenance`, and `advanced`. The first five are regular-user entry points; `plots` retains the core report graphics and `advanced` contains the complete canonical result unchanged. Historical direct access to canonical fields (for example `result$optimization`) remains supported.

Direct `semantica_full_pipeline()` results remain the complete advanced object. `semantica_view(result, view = "advanced", section = "structural")` provides the same grouped navigation for either interface without recomputation. Use `semantica_items()` for the normalized wording table and `summary(result)` for the detailed diagnostics and evidence boundary.

# 6. Plot and save

```{r eval=FALSE}
plot(result)
result$plots$plot_fitness_evolution
result$plots$plot_esem_before
result$plots$plot_esem_after
result$plots$plot_pfa_diagnostics
plot(result, which = "aco")
plot(result, which = c("esem", "pfa"))
plots <- semantica_plot_all(result, progress = FALSE)
attr(plots, "semantica_plot_manifest")

semantica_export(result, "academic_self_regulation")
semantica_save_bundle(result, "academic_self_regulation_bundle.rds")
```

`semantica_export()` is readable report/interchange output. It is not the inverse of `semantica_reload()`, which reads the component-level optimizer-interchange format. `semantica_save_bundle()` is the provenance-preserving analysis bundle.

# 7. When to use `semantica_full_pipeline()`

Use `semantica_full_pipeline()` for participant-response validation, custom
quality policies, detailed ESEM/PFA controls, specialized compute settings, or
additional diagnostics.

The default workflow uses the conservative content-alignment guard and the
relative-conservative semantic objective. `semantica_quality_config()` exposes
these and related quality settings.

# Existing items and participant data

These are common supported workflows, but they begin at a different point in
the pipeline. See:

```r
vignette("semantica-user-workflows", package = "SEMANTICA")
```

# Interpretation boundary

Sample-free semantic/PFA/ESEM/DFI results are representation-derived proxy
diagnostics. They are useful for pre-data screening, but they are not
participant-response evidence of reliability or validity. After item selection,
review the content substantively and collect appropriate response data.
