---
title: "SEMANTICA Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEMANTICA Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Start here

Before rerunning an expensive analysis:

```{r eval=FALSE}
semantica_list_backends()
semantica_check_setup(llm = your_llm_config, probe = TRUE)
semantica_run_plan(...)
```

# Missing API key

Store cloud credentials in `.Renviron` and restart R. Common variables are
`OPENAI_API_KEY`, `ANTHROPIC_API_KEY`, `GROQ_API_KEY`, and `HF_TOKEN`.

Do not place credentials in saved SEMANTICA configuration files or public
scripts when an environment variable is sufficient.

# Anthropic or Groq says embeddings are unavailable

This is expected for the built-in registry. Configure a separate embedding
backend:

```{r eval=FALSE}
llm_cfg <- semantica_llm_config(
  backend = "anthropic",
  embed_backend = "openai"
)
```

# Ollama/local server is unreachable

Verify that the server process is running and that the model is installed.
Then use:

```{r eval=FALSE}
semantica_check_setup(llm = "ollama", probe = TRUE)
```

A preflight warning is diagnostic; the actual request remains authoritative.

# Model not found

Model identifiers are provider/local-server specific. Use the provider's model
registry or local model list, then rerun `semantica_check_setup(..., probe =
TRUE)`.

# Existing item pool fails

Check that the table has unique, non-empty `item_id`, `factor`, and `item_text`
columns. For external embeddings, row IDs must exactly match item IDs.

# Too many items/API work were requested

Remember:

- `pool_items` is retained candidates **per factor**;
- `overgenerate` multiplies the approximate raw target;
- facets divide the factor-level retained target rather than multiplying it.

Use `semantica_run_plan()` before the next run.

# Console output is too verbose

Use:

```{r eval=FALSE}
semantica_run(..., progress = "normal")   # concise default
semantica_run(..., progress = "quiet")
semantica_run(..., progress = "detailed")
```

# An interrupted parallel run left resources uncertain

Use:

```{r eval=FALSE}
semantica_reset_resources()
```

Then rerun with `workers = "auto"`, an explicit worker limit, or
`workers = "serial"`.

# Cache behavior is confusing

Inspect the cache before changing it:

```{r eval=FALSE}
semantica_cache_info()
semantica_cache_info(result)
```

To bypass the cache, set `embedding_cache = FALSE` in
`semantica_llm_config()`. To delete cached embeddings, explicit confirmation is
required:

```{r eval=FALSE}
semantica_clear_cache(confirm = TRUE)
```

# Python/Hugging Face backend does not initialize

Use the supported environment helpers first:

```{r eval=FALSE}
semantica_setup_conda()
semantica_activate_conda()
semantica_check_setup(llm = "python_hf", probe = TRUE)
```

Only move to manual device-map/CUDA settings after the base environment is
working.

# A plot fails but the run succeeded

Plot generation is post-analysis presentation. `semantica_plot_all()` now
reports how many plot objects succeeded and lists optional failures. A failed
plot does not retroactively invalidate the completed optimizer result.

# How do I know what settings were really used?

```{r eval=FALSE}
semantica_config(result)
semantica_models(result)
```

These access the sanitized resolved configuration/provenance recorded by the
pipeline.

# What should I do after a successful sample-free run?

1. inspect `semantica_items(result)`;
2. review the wording and content coverage substantively;
3. inspect `semantica_evidence_status(result)` and `summary(result)`;
4. preserve the bundle;
5. collect participant responses;
6. perform appropriate response-based psychometric validation before making
   reliability/validity claims.
